---
title: "Final Report"
author: "Radiah Khan '28 | Faculty Advisor: Dr. Kementari Whitcher"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: true
    fig-cap-location: top
editor: source
code-fold: true
execute: 
  echo: true
  warning: false
  error: false
  message: false
knitr:
  opts_chunk: 
    message: false
bibliography: references.bib

---

```{r, include=FALSE}
library(dplyr)
library(tidyverse)
library(dae)
library(Stat2Data)
library(dae)
library(Stat2Data)
library(datarium)
library(ggplot2)
library(ggpubr)
library(rstatix)  
library(tmap)
library(ggthemes)
library(cartogram)
library(ggrepel)
library(binom)
library(car)
library(exactci)
library(PropCIs)
library(gam)
library(mfx)
library(pROC)
library(flextable)
library(emmeans)
library(treemapify)
library(flextable)
library(knitr)
library(kableExtra)
library(gt)
library(dplyr)
library(tidyr)
library(highcharter)
library(webshot2)
library(readxl)
library(gridExtra)


```

```{r, include=FALSE}
table <- data.frame(
  Credit = c("Food and Beverage Purchasing", "Sustainable Dining"),
  Point = c("6", "2"))
```

```{r tab:scoresystem, echo=FALSE}
kable(table, 
      col.names = c("Credit", "Point"), 
      align = c("c", "c"),
      caption = "Credit system for dining impact area (Version 2.2)") |> 
  kable_styling(full_width = TRUE, position = "center")

```




```{r, include=FALSE}
table2 <- data.frame(
 Budget = c("1 million - 4.9 million", "5 million - 9.9 million", "10 million or more"),

  Class = c("1", "2", "3"))
```

```{r tbl-budgetclass, echo=FALSE}
kable(table2, 
      col.names = c("Budget", "Class"), 
      align = c("c", "c"),
      caption = " ") |> 
  kable_styling(full_width = TRUE, position = "center")




```

## Score Ratio {#sec-score-ratio}

The points earned by institutions in the impact area of food and dining had two score scales : out of 6 and out of 8; depending on the year. To keep the calculation consistent over all years; we calculated the proportion of the score to the total possible as a score ratio.

For example @tbl-score-ratio shows how we calculated the score ratio for these two cases. 
```{r tbl-score-ratio, echo=FALSE}
table3 <- data.frame(
 ` Score given by STARS` = c ("3/6", "4/8"),
 `Score Ratio` = c("0.5", "0.5")
)
kable(table3, caption = "")
```





```{r, include=FALSE}
library(readxl)
library(binom)
library(exactci)
library(PropCIs)
library(epitools)
library(vcd) 
library(vcdExtra)
library(tidyverse)
library(car)
library(flextable)
```

```{r, include=FALSE}
tests_smith <- data.frame(
  Variables = c("Category, Local", "Local_2017, Cost_2017","Local, Cost"),
  Test = c("Chi Square", "Logistic Regression", "Logistic Regression")
)
```

```{r include=FALSE}
flextable(tests_smith)|>
  set_caption("Statistical Tests Conducted in Smith College Data")|>
  autofit()|>
  theme_apa()
```

### Defining the variables used

The variables used for analyzing Smith College data are shown in @tbl-variable

```{r tbl-variable, echo=FALSE}
smith_variables <- data.frame(
  Variable = c("Local_2017", "Local_2018", "Cost_2017", "Cost_2018", "Category", "local_proportion", "count_yes", "count_no"),
  Type = c("Binary", "Binary", "Numerical", "Numerical", "Categorical", "Numerical","Numerical", "Numerical"),
  Interpretation = c(
    "Coded as 1(local) and 0 (non-local) in 2017",
    "Coded as 1(local) and 0 (non-local) in 2018",
    "Price of each food item in 2017",
    "Price of each food item in 2018",
    "STARS standard categories of food that calculates to total food and beverage expenditure", 
    "The proportion of food categorized as local",
    "The count of food categorized as local",
    "The count of food categorized as non-local"
  )
)

kable(smith_variables, caption = " ")

```

```{r, include= FALSE}
library(readxl)
logistical_regression_2017 <-  read_excel("~/Downloads/Smith_College_RFC_assessment-2017-10and_2018-02 (2)-2.xlsx", sheet = "Logistic-regression")



colnames(logistical_regression_2017) <- c("Local_2017", "Cost_2017", "Local_2018", "Cost_2018")
logistical_regression_2017
```

```{r, include=FALSE}
library(dplyr)
logistical_regression_2017 <- logistical_regression_2017 |>
dplyr::select( "Local_2017"    ,  "Cost_2017"            )

dplyr::select(logistical_regression_2017, Local_2017, Cost_2017)

```



```{r fig-logistic-regression-2017, echo=FALSE}

plot(jitter(logistical_regression_2017$Local_2017, 0.08) ~ logistical_regression_2017$Cost_2017,
     main = "Scatterplot of Logistical Regression Analysis (2017)",
     xlab = "Cost",
     ylab = "Local Status (Jittered)")

library(gam) # for fitting generalized additive models
gam.fit <-gam(Local_2017 ~ s(Cost_2017), family=binomial, data=logistical_regression_2017) # s = smooth funct.
curve(predict(gam.fit, data.frame(Cost_2017=x), type="resp"), add=TRUE) # plots generalized additive model smoothing fit
fit <-glm(Local_2017 ~ Cost_2017, family=binomial, data=logistical_regression_2017) # link=logit is default
curve(predict(fit, data.frame(Cost_2017=x), type="resp"), add=TRUE) # logistic regression fit is added to the plot > summary(fit)

```{r, include = FALSE}
model_data <- model.frame(fit2017)

```


```{r fig-roc-2017, echo=FALSE}
fit2017 <-glm(Local_2017 ~ Cost_2017, family=binomial, data=logistical_regression_2017)
rocplot <- roc(response = model_data$Local_2017, predictor = fitted(fit2017))
plot.roc(rocplot, legacy.axes=TRUE) # Specficity on x axis if legacy.axes=F
auc_value <- auc(rocplot)

text(x = 0.2, y = 0.2, labels = paste("AUC =", round(auc_value, 4)), adj = 0, cex = 1.2, col = "black")
```

#### 2018

```{r, include=FALSE}
setwd("~/Downloads") 
logistic_regression_2018 <-readxl::read_excel("Logistic Regression-2018.xlsx")
logistic_regression_2018
```
We will conduct the same test with 2018 data as well to see if the association is statistically significant or not. 
```{r, echo=FALSE}
fit2018<-glm(Local ~Cost,family=binomial(link=logit),data=logistic_regression_2018)
summary(fit2018)

```


```{r fig-lr-2018 , echo=FALSE}
lr_plot2 <-
plot(jitter(Local, 0.08) ~ Cost, data=logistic_regression_2018,
     main = "Scatterplot of Logistical Regression Analysis (2018)",
     xlab = "Cost",
     ylab = "Local Status (Jittered)")

library(gam)
gam.fit <- gam(Local ~ s(Cost), family=binomial, data=logistic_regression_2018)
curve(predict(gam.fit, data.frame(Cost = x), type="resp"), add=TRUE)

fit <- glm(Local ~ Cost, family=binomial, data=logistic_regression_2018)
curve(predict(fit, data.frame(Cost = x), type="resp"), add=TRUE)
auc_value <- auc(rocplot)

text(x = 7000, y = 0.2, labels = paste("AUC =", round(auc_value, 4)), adj = 0, cex = 1, col = "black")


lr_plot2


```

The @fig-lr-2018 shows that more values cluster around lower cost.

```{r, include=FALSE}
model_data_2 <- model.frame(fit2018)
```

```{r, echo=FALSE}
fit2018 <-glm(Local ~ Cost, family=binomial, data=logistic_regression_2018)
rocplot <- roc(response = model_data_2$Local, predictor = fitted(fit2018))
plot.roc(rocplot, legacy.axes=TRUE) # Specficity on x axis if legacy.axes=F
auc(rocplot) # auc = area under ROC curve = concordance index
```

- **Null Hypothesis**  ($H_0$): The two variables are independent.

- **Alternative Hypothesis** ($H_1$): The two variables are not independent.

We start by extracting the Category column and Local column from the 2017 and 2018 data set. Then, we grouped the data by the category. Furthermore, we calculated the amount of local item in that category and the non-local item amount as well.

### 2017

Firstly, we analyze 2017 data with these two variables: Local and Category

```{r, include=FALSE}
library(readxl)

one_way_2017 <- read_excel(path = "~/Downloads/Smith_College_RFC_assessment-2017-10and_2018-02 (2)-2.xlsx",sheet = "10-17"
)
```

```{r, include=FALSE}
one_way_2018 <- read_excel(path = "~/Downloads/Smith_College_RFC_assessment-2017-10and_2018-02 (2)-2.xlsx",sheet = "2-18"
)
```

```{r, include=FALSE}
one_way_2017_tibble_1 <- one_way_2017 |>
  dplyr::select(Category, Local) |>
   dplyr::group_by(Category, Local) |>
   dplyr::summarise(count = n(), .groups = "drop") |>
   dplyr::group_by(Category) |>
   dplyr::mutate(
    proportion = count / sum(count)
  ) |>
  pivot_wider(names_from = Local, values_from = c(count, proportion)) |>
  mutate(
    total_count = count_yes + count_no
  ) |>
  filter(!is.na(count_yes))

kable(one_way_2017_tibble_1, caption = "")

one_way_2017_tibble_1
```

```{r, include=FALSE}
one_way_2018_tibble_2<- one_way_2018 |>
  dplyr::select(Category, Local) |>
   dplyr::group_by(Category, Local) |>
   dplyr::summarise(count = n())|>
   dplyr::group_by(Category) |>
   dplyr::mutate(proportion = count / sum(count))|>
  pivot_wider(names_from = Local, values_from = c(count,proportion))|>
 filter(!is.na(count_yes)) |>
mutate(total_count = count_yes + count_no)
one_way_2018_tibble_2
```

```{r, include=FALSE}
one_way_2017_tibble <- one_way_2017_tibble_1 |> 
  filter(!is.na(count_yes))
```

```{r, include=FALSE}
one_way_2018_tibble <- one_way_2018_tibble_2 |> 
  filter(!is.na(count_yes))
```

```{r, include=FALSE}
two_way_2017<- one_way_2017_tibble |>
 dplyr::filter(Category != "eggs", Category != "teacoffee") |>

  dplyr::select(Category, count_yes,count_no) 
```

```{r, include = FALSE}
two_way_2017_table <- two_way_2017 |>
  tibble::column_to_rownames(var = "Category") |>
  as.matrix() |>
  as.table()
print(two_way_2017_table)
```

```{r, echo=  FALSE}
chisq.test(two_way_2017_table, correct = FALSE)
```

The p value is extremely close to 0; so it is less than the significance level 0.05.

Hence we can reject the null hypothesis. It can be concluded that the two variables are not independent.If they are not independent, then some categories have a higher proportion of local than others.

```{r fig-count-2017, fig.cap="", echo=FALSE}
ggplot(data = two_way_2017,
       aes (x = fct_reorder(Category, count_yes, .desc = TRUE),
            y = count_yes))+
   geom_col(fill = "royalblue") +
  labs(
    title = "Count of Local Products Purchased (2017)",
    x = " ",
    y = " "
  ) +
  theme_minimal() +
theme(plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major=  element_blank(),
      panel.grid.minor = element_blank())
      
```

From @fig-count-2017 we can see that counts of product categories that are local, we can see that produce, grocery, meat, dairy has higher counts of local sourcing.

### 2018
Then, we analyze 2018 data with the same variables to see if it is not independent in 2018 as well.


```{r, include=FALSE}
two_way_2018<- one_way_2018_tibble |>
 dplyr:: select(Category, count_yes,count_no) |>
  dplyr::filter(!is.na(count_yes))
```

```{r, include=FALSE}
two_way_2018_table <- two_way_2018 |>
  tibble::column_to_rownames(var = "Category") |>
  as.matrix() |>
  as.table()
print (two_way_2018_table)

```

After conducting the chi square test in 2018, we see the following results.

```{r, echo=FALSE}
chisq.test(two_way_2018_table, correct = FALSE)
```

The p value is extremely close to 0; so it is less than the significance level 0.05. Hence we can reject the null hypothesis. It can be concluded that the two variables are not independent.If they are not independent, then some categories have a higher proportion of local than others.

Now that we know that for both of the years some categories have higher proportion of being local than the other ones, we will plot this.

```{r fig-count-2018, fig.cap="", echo=FALSE}
ggplot(data = two_way_2018,
       aes (x = fct_reorder(Category, count_yes, .desc = TRUE),
            y = count_yes))+
   geom_col(fill = "royalblue") +
  labs(
    title = "Count of Local Products Purchased (2018)",
    x = " ",
    y = " "
  ) +
  theme_minimal() +
theme(plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major=  element_blank(),
      panel.grid.minor = element_blank())
      

```

From the @fig-count-2018, we can see that our test results reflect on it. Dairy has the highest count of local purchases, followed by produce, grocery, meat and the other categories.

## Category and Difference in Proportion from year 2017-2018

Now that we know some categories have a higher count than the other. We want to test if the proportion of the categories which were local is significantly different between the year 2017 and 2018.

```{r, include = FALSE}
two_prop_test <- function(yes_2017, yes_2018, total_2017, total_2018) {
test_result<-prop.test(x=c(yes_2017, yes_2018),n=c(total_2017, total_2018), correct = FALSE)
print(test_result)
}
```

```{r, include = FALSE}
test_result_baked <- two_prop_test(18,2,172,278)
```

```{r, include = FALSE}
test_result_bev <- two_prop_test(6,8,154,209)
```

```{r, include = FALSE}
test_result_dairy <- two_prop_test(53,411,207,545)
```

```{r, include = FALSE}
test_result_grocery <- two_prop_test(79,168,1019,1023)
```

```{r, include= FALSE}
test_result_meat <- two_prop_test(58,37,143,120)

```

```{r, include = FALSE}
test_result_poultry <- two_prop_test(4,4,40,39)
```

```{r, include = FALSE}
test_result_produce <- two_prop_test(115,188,495,533)
```

After conducting the two proportion test on each category here are the results in @tbl-two-prop

```{r tbl-two-prop, echo=FALSE}
test_results_two_prop_test <- data.frame(
  
  Categories = c("baked", "beverages", "dairy", "grocery", "meat", "poultry", "produce"), 
  `p value` = c("less than 0.05", "not significant", "less than 0.05", 
              "less than 0.05", "not significant", "not significant", 
              "less than 0.05")
)

kable(test_results_two_prop_test, caption = " ")





```

The significance level is 0.05. From the table we can see baked, dairy, grocery and produce has p \< 0.05, hence the proportion of these categories which were local is significantly different between the year 2017 and 2018.

Now that we know the change was significant, we will plot these changes. 

In @fig-local-slopegraph we can see that dairy had the highest increase in locally sourced proportion from the year 2017 to 2018. Followed by produce and grocery. Baked had a decrease in proportion.   

```{r, include= FALSE}
combined_evoluation <- data.frame(
  Year = rep(c("2017", "2018"), times = 7), 
  Category = rep(c("baked", "beverages", "dairy", "grocery", "meat", "poultry", "produce"), each = 2),
  local_proportion = c(0.10465116,  0.007194245, 0.03896104,0.038277512,  0.25603865, 0.754128440,  0.07752699, 0.164222874, 0.40559441, 0.308333333, 0.10000000, 0.102564103,  0.23232323, 0.339963834 )
)|>
  mutate(local_proportion = round(local_proportion,2))

combined_evoluation


```
```{r fig-local-slopegraph, fig.cap=" ", echo= FALSE}
library(CGPfunctions)

newggslopegraph(dataframe = combined_evoluation,
                Times = Year,
                Measurement = local_proportion,
                Grouping = Category,
                Title = "Local Produce Evolution at Smith College (2017 to 2018)",
                SubTitle = "",
                Caption = "By CEEDS Data",
                DataLabelPadding = 0.1,
                
                DataLabelLineSize = 0.0123,
                DataLabelFillColor = "lightblue",
                ThemeChoice = "void")


```

## Cash Flow at Smith College Over the Year of 2017-18

We know that the proportion in each category and their local-non local status is not independent. There is statistical association between these two variables.

Now, we will visualize how the cash flow changed over the year (2017-2018) for the same categories.

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false 

Book3_6 <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Smith College Data/Book3-6.xlsx")

```
```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false 
Book3_6_2 <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Smith College Data/Book3-6-2.xlsx")
```
```{r fig-cash, fig.cap = "", fig.width=8, fig.height=6, echo=FALSE}
 
#| error: true
cost_distribution_plot  <- hchart(Book3_6
  , 
  "sankey",
  hcaes(from = From, to = To, weight = Value, color = Color)
) |>
 hc_title(
    text = "Cost Distribution per Category over 2017-18",
    align = "center",                     
    style = list(fontWeight = "bold")    
  ) |>
  hc_add_theme(hc_theme_ggplot2())
cost_distribution_plot

```

```{r, fig-amount, fig.cap="", fig.width=8, fig.height=6, echo=FALSE}
amount_distribution_plot <- hchart(Book3_6_2
  , 
  "sankey",
  hcaes(from = From, to = To, weight = Value, color = Color)
) |>
 hc_title(
    text = "Amount Distribution per Category over 2017-18",
    align = "center",                     
    style = list(fontWeight = "bold")    
  ) |>
  hc_add_theme(hc_theme_ggplot2())
amount_distribution_plot


```


```{r, include=FALSE, error = TRUE}
group_total <- data.frame(
  Group = c("Real", "Sustainable", "Local", "Fair", "Humane"),
  Value = c(1511918.89, 441283.02, 892277.78, 116531.66, 180724.49)
)
group_total
```

```{r fig-tree-2023, fig.cap = "",fig.width=8, fig.height=6, echo=FALSE}
plot <- ggplot(group_total, aes(area = Value, fill = Value, label = Group)) +
  geom_treemap() +
  geom_treemap_text(colour = "black",
                    place = "centre",
                    size = 16,
                    grow = FALSE) +
  scale_fill_gradientn(
    name = "Total Spending (in Dollars",
    colors = c("yellow", "lightblue", "blue"),
    breaks = scales::pretty_breaks(n = 4),
    labels = scales::label_comma()
  ) +
  labs(title = "Spending of each food produce type at Smith College in 2023") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.text = element_text(size = 6.5),
    legend.title = element_text(size = 7)
  )

plot

```


```{r, include=FALSE}
sustainable_five <- data.frame(
  College = c("Smith College", "UMass Amherst", "Hampshire College"),
  Percentage = c("23.38%", "23%", "33.02%")
)
sustainable_five

```

```{r, echo= FALSE}
ggplot(data = sustainable_five, 
        aes(x = fct_reorder(College, Percentage, .desc = TRUE), 
             y = Percentage, fill = College)) +
  geom_col(show.legend = FALSE, color = "black", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Percentage of Sustainable Produce",
    x = "",
    y = " "
  ) +
 theme_minimal() |>
  theme(
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
    
  ) 
  

```

```{r, include=FALSE}
Smith_College_2018 <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Smith College Data/Smith_College_RFC_assessment-2017-10and_2018-02 (2)-3.xlsx", 
                                 sheet = "2-18") |>
  dplyr::select("Local", "Fair", "Ecological", "Humane", "Cost") |>
  mutate("Sustainable" = Local == "yes" | Fair == "yes" | Ecological == "yes" | Humane == "yes")|>
  dplyr::select(Sustainable, Cost) |>
  group_by(Sustainable)|>
  summarize(total = sum(Cost, na.arm = TRUE))

Smith_College_2018
```

### Locally Produced Percentage Comparison: Smith College vs Hampshire vs Mount Holyoke College vs Amherst College

For Smith College, we only selected the local variable. The products that are local were calculated and then divided it by total cost of all local-non local products.

For @umass, they signed the Real Food Challenge Commitment and identified food that were third party verified(organic or humane)/ local-community sources.

For Hampshire College, they did not have categories specified. However they provided a percentage of food that were local-community based.

```{r, include=FALSE}

smith_18 <- one_way_2018 |>
 dplyr:: select(Cost, Local)|>
group_by(Local)|>
  summarize(total = sum(Cost, na.arm = TRUE))
 smith_18 
 
  
```

```{r, include=FALSE}
Amherst_College_2019 <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Five College Data/Amherst College 2019.xlsx")
View(Amherst_College_2019)
```

```{r, include=FALSE}
library(janitor)
amherst2019 <- Amherst_College_2019 |>
  janitor::clean_names()

amherst2019
colnames(amherst2019)
```

```{r, include=FALSE}
amherst_data <- amherst2019 |>
  dplyr:: select("amount_spent", "local_community_based_2"   )|>
  group_by(local_community_based_2)|>
  summarize(total_cost = sum(amount_spent, na.rm = TRUE))
amherst_data
```

```{r, include=FALSE}

smith_2018 <- one_way_2018|>
  dplyr:: select(Local, Cost)|>
  group_by(Local)|>
  summarize(total_cost = sum(Cost, na.rm = TRUE))
smith_2018
```

```{r, include=FALSE}
Mount_Holyoke_College <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Five College Data/Mount Holyoke College.xlsx", sheet = "Sheet3")
View(Mount_Holyoke_College)

```

```{r,include=FALSE}

combined_five_college <- data.frame(
  `College Name` = c("Smith College (2018)", "Mount Holyoke College (2018)", "Hampshire College (2021)", "Amherst College(2019)"),
  `Local Percentage` = c(13.48, 9.51 ,26, 58),
  Year = c(2018, 2018, 2021, 2019)
)
combined_five_college

```

```{r fig.width=8, fig.height=6, echo=FALSE}
ggplot(data = combined_five_college, 
        aes(x = fct_reorder(College.Name, Local.Percentage, .desc = TRUE), 
             y = Local.Percentage, fill = College.Name)) +
  geom_col(show.legend = FALSE, color = "black", width = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Percentage of Locally Produced Purhcases",
    x = "",
    y = "Local Percentage (%)"
  ) +

  theme_minimal(base_size = 10) |>
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
    
  ) 

```

## All College Categories Combined

In this section, we will analyze all of the institutions across the United States. The analysis will be of three types. 
-  Comparison Across Years
-  Comparison Between Categories (Flagship and Private)
-  Correlation between variables used in @sec-data-variable

### Score Ratio- Budget Class

We are going to analyze the three budget classes that we classified in the institutions across institutions in @tbl-budgetclass, the score ratio that we determined for the institutions @tbl-score-ratio and year (@sec-year) with a two way ANOVA test. 

The two factor in this ANOVA test is `Budget Class` and `Year` where the dependent variable is the score ratio. 

- **Null Hypothesis 1** : Budget Class has no effect on the score ratio. 
- **Null Hypothesis 2**: Year has no effect on the score ratio.
- **Null Hypothesis 3**: There is no interaction effect between the two factors.

- **Alternative Hypothesis 1**: Budget Class has a significant effect on the score ratio.
- **Alternative Hypothesis 2**: Year has a significant effect on score ratio.
- **Alternative Hypothesis 3**: There is an interaction effect between two factors

```{r, include=FALSE}
library(readxl)
combined <- Combined_Database_2 <- read_excel("~/Downloads/SURF 2025 _ SDS _ Dr. Whitcher & Radiah-3/Combined Data/Combined Database -2.xlsx", sheet = "Combined Data(Usable)")
View(Combined_Database_2)
```

```{r, include= FALSE}
combined_budget_score_1 <- combined |>
dplyr::select(`Budget Class Year 1`, `Score ratio Year 1`, `Score ratio Year 2`)|>
filter(!is.na(`Budget Class Year 1`), !is.na(`Score ratio Year 1`),!is.na(`Score ratio Year 2`) 
  )
```

```{r, include=FALSE}
combined_budget_score <-combined_budget_score_1 |>
 pivot_longer(
    cols = c("Score ratio Year 1", "Score ratio Year 2"),
    names_to = "Year",                        
    values_to = "value"                       
  )
```

```{r, include=FALSE}
library(palmerpenguins)
library(knitr)
```

```{r, include=FALSE}
summary_table_budget_score <- summary(combined_budget_score)
```

```{r, include=FALSE}
kable(summary_table_budget_score, caption = "Summary of Budget Class and Sustainibility Score Ratio Over The Years")
```

```{r, include=FALSE}
combined_budget_score <- combined_budget_score |>
  mutate(Year = factor(Year)) |>
  mutate(`Budget Class Year 1`= factor(`Budget Class Year 1`))|>
   rename( `Budget Class`= `Budget Class Year 1`)
combined_budget_score
```

```{r, include= FALSE}
model_1 <- lm(value ~  `Budget Class`* Year,
  data = combined_budget_score
)
```

```{r, echo=FALSE}
anova(model_1)
```

##### Result

The significance level is 0.05. So since the p values are greater than 0.05, none of the factors are significant. Budget level and year do not make a difference ratio of score that is based on agglomerated sustainability of schools.

## Total food spend, Score Ratio

```{r, include=FALSE}
combined_total_score <- combined |>
  dplyr:: select(
    `Score ratio Year 1`, 
    `Score ratio Year 2`, 
    `Total food and beverage expenditures  Year 1`, 
    `Total food and beverage expenditures Year 2`
  ) |>
dplyr::rename(
    `Year 1_S` = `Score ratio Year 1`,
    `Year 2_S` = `Score ratio Year 2`,
    `Year 1_T` = `Total food and beverage expenditures  Year 1`,
    `Year 2_T` = `Total food and beverage expenditures Year 2`
  ) |>
  
dplyr::mutate(
    `Year 1_T` = as.numeric(`Year 1_T`),
    `Year 2_T` = as.numeric(`Year 2_T`)
  )

```

```{r, include=FALSE}
combined_total_score_long <- combined_total_score |>
  pivot_longer(
    cols = c(`Year 1_S`, `Year 2_S`),
    names_to = "Score ratio year",                        
    values_to = "Score Ratio"                       
  ) |>
  pivot_longer(
    cols = c(`Year 1_T`, `Year 2_T`),
    names_to = "Budget Year",
    values_to = "Total Budget"
  )
```

```{r, include=FALSE}
combined_total_score_cleaned <- combined_total_score_long |>
  dplyr:: select(`Score Ratio`, `Total Budget`)
combined_total_score_cleaned

```

```{r, include=FALSE}
combined_budget_score_filtered <- combined_total_score_cleaned |>
  filter(!is.na(`Score Ratio`)) |>
    filter(!is.na(`Total Budget`)) 
combined_budget_score_filtered

```


We performed a linear regression analysis here where the Total Budget is the independent variable and the Score Ratio is the dependent variable. 

- **Null Hypothesis**: $H_0: \beta = 0$; There is no association between the independent and dependent variable 
-  **Alternative Hypothesis**: $H_0: \beta \neq 0$; There is statistically significant association between the independent and dependent variable.

```{r fig-scatter-budget, fig.cap="", echo=FALSE}
ggplot(
  data = combined_budget_score_filtered,
  mapping = aes(
    x = `Total Budget`,
    y = `Score Ratio`
  )
) + 
  geom_point()+
  geom_smooth(method= "lm", se= FALSE, color = "red")
```

From the @fig-scatter-budget, we can see there is a weak positive relationship between the two variable.

```{r,include=FALSE}
library(janitor)
```

```{r, include=FALSE}
cleaned_scores <- clean_names(combined_budget_score_filtered)

glimpse(combined_budget_score_filtered)
  
```

The resulting model shows that there is a weak but statistically significant positive association between the two variable. (t = 2.449 with 114 degrees of freedom, p \<0.05) 

 **Linear Equation**:  
  Score Ratio = 0.2092 + 5.047 * 10^-9 * Total Budget  
  R-squared: 4.2%

```{r, echo=FALSE}
score_budget_model =lm(score_ratio ~ total_budget, data= cleaned_scores)
summary(score_budget_model)
```

```{r, echo=FALSE}
anova(score_budget_model)
```

So we reject the null hypothesis and conclude that $\beta \neq 0$, which means there is a statistically significant association between the two variables.

```{r, include=FALSE}
score_budget_model.stdres = rstandard(score_budget_model)
cleaned_score_budget <- data.frame(cleaned_scores, score_budget_model.stdres)
glimpse(cleaned_score_budget)
```

```{r, include=FALSE}
ggplot(
  data = cleaned_score_budget,
  mapping = aes(
    x = total_budget,
    y = score_budget_model.stdres
  )
) + 
  theme_classic()+
  geom_point()
```

To check the normality of the residual plots for this analysis, we used a Q-Q plot to visualize it.

```{r, echo=FALSE}
qqnorm( score_budget_model.stdres,
       ylab="Standardized Residuals",
       xlab="Normal Scores")
qqline( score_budget_model.stdres)
```

In here we can see that it performs reasonably well along most of the data with deviations from the expected pattern at the high end.

Now, what does it mean in our context?

The more an institution spent on its dining sector (based on the STARS categories), the higher its score ratio. This score indicates better performance in sustainable food procurement.

It seems to be the case that institutions that have a larger student body and higher funding spend more money on food and beverages expenditure, eventually gets a higher score ratio.

It is also important to reiterate that the score ratio criteria differ from year to year. The earlier year (Version 2.2 STARS) relied on 2 categories: Sustainable Dining, Food and Beverage Purchasing. Whereas recent years (Version 3.0.1 STARS) relied on two categories: Percentage of food and beverage spend that meets sustainability criteria and percentage of dining service spend with social impact suppliers. Converting their score to a ratio based on the total score possible made comparisons across years and institutions possible.

## Score ratio and Category of Insitutions 
Now we will conduct an analysis on the score ratio and the categories of institutions (Flagship and Private) to see if their mean score ratio differs or not. 

### Year 1

In this part, we compare the score ratios from two different category of institutions(flagship & private) for the first year of data of colleges. 

-  **The null hypothesis** ($H_0$): The means of score ratio of the categories of institutions are equal.\\
- **The alternative hypothesis** ($H_1$): The means of score ratio of the categories of institutions are not equal.

The significance level is $0.05$.For visualizing the distribution of the data, we use a box plot before conducting the independent sample t-test.

```{r, include = FALSE}
score_school_category<-combined |>
  dplyr::select(`Score ratio Year 1`, Status)|>
  dplyr::rename(score_ratio = `Score ratio Year 1`,
         category = Status) |>
  dplyr::mutate(category = factor(category))  |>
  dplyr::mutate(score_ratio = log(score_ratio))
score_school_category
```

```{r, include=FALSE}
score_school_category_cleaned <- score_school_category |>
  filter(
    (category == "Flagship" & score_ratio <= 0.5) |
    (category == "Private" & score_ratio <= 0.75)
    
  )
score_school_category_cleaned
```

```{r fig-score-category-year1, fig.cap=" ", echo=FALSE}
ggplot(data = score_school_category_cleaned, mapping = aes(x = score_ratio, y = category, fill = category)) +
theme_classic() +
  theme(
plot.title = element_text(hjust = 0.5, face = "bold"),
 axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  labs(
    title = "Distribution of Score Ratios by School Category",
    x = "Score Ratio",
    y = "Value",
    fill = "School Category",
    caption = "Year 1"
  )+
 
  geom_boxplot() +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black")+
  coord_flip()
```

In the @fig-score-category-year1, we removed outliers to improve normality. The median value is different in two categories. The private institutions show more dispersion compared to flagship institutions.

Before conducting the t-test, we conduct a Levene test.

```{r, echo=FALSE}
 score_school_category_cleaned |>
  levene_test(score_ratio ~ category)
```

We can see that the p \> 0.05 (significance level), so it is not statistically significant. Hence we do a t-test assuming equal variances.

```{r, include=FALSE}
library(rstatix)
```

```{r, echo=FALSE}
t.test(score_ratio ~ category, data = score_school_category_cleaned, var.equal= TRUE)
```

While there is a slight difference in the means of Flagship and Private, it is not statistically significant. Flagship has lower mean than Private. The p-value is 0.4111 which is not less than p value of 0.05. So we fail to reject the null hypothesis that there is no significant difference between score ratio in different college categories.

To put this in context,there appears to be no significant difference in the mean score ratio between flagship and private schools. From our previous test we saw that there is a positive significant association between dining expenditure and score ratio. We assumed that the category of colleges might play a role in the association. From the test we know that it is not significant.

## Year 2

Now we conduct the same process for data of year 2 of the reporting institutions.

```{r, include=FALSE}
score_school_category_2 <-combined |>
  dplyr:: select(`Score ratio Year 2`, Status)|>
  rename(score_ratio = `Score ratio Year 2`,
         category = Status) |>
  mutate(category = factor(category))  |>
  #mutate(score_ratio = log(score_ratio))|>
  filter(!is.na(score_ratio)) |>
  filter(
    (category == "Flagship") |
    (category == "Private" & score_ratio <= 0.6)
  )
  
score_school_category_2
```

```{r fig-score-category-year2, fig.cap=" ", echo=FALSE}
ggplot(
  data = score_school_category_2,
  mapping = aes(
    x = score_ratio,
    fill = category
    )
  ) +
  theme_classic() +
  theme(
plot.title = element_text(hjust = 0.5, face = "bold"),
 axis.text.x = element_text(angle = 45, hjust = 1)
  )+
  labs(
    title = "Distribution of Score Ratios by School Category",
    x = "Score Ratio",
    y = "Value",
    fill = "School Category",
    caption = "Year 2"
  )+
 
  geom_boxplot()
```

In the @fig-score-category-year2, we removed outliers to improve normality. The median value is slightly different in two categories. The private box is slightly wider than flagship box.

From the levene test we can see that p\>0.05, so we use simple t-test assuming variance is equal.

```{r, echo=FALSE}
score_school_category_2 |>
  levene_test(score_ratio ~ category)
```
From the levene test we can see that p\>0.05, so we use simple t-test assuming variance is equal.

```{r, echo=FALSE}
t.test(score_ratio ~ category, data = score_school_category_2, var.equal=TRUE)
```

The p-value is 0.84 which is not less than p value of 0.05. So we fail to reject the null hypothesis that there is no significant difference between score ratio in different college categories in year 2 as well.

## Ethically Produced Rate and Budget

We performed a linear regression analysis here where the Total Budget is the independent variable and the Ethically Produced Rate is the dependent variable. 

- **Null Hypothesis** $H_0$: $\beta = 0$ : There is no association between the independent and dependent variable 
- **Alternative Hypothesis** $H_0$: $\beta \neq 0$ : There is statistically significant association between the independent and dependent variable.

```{r, include=FALSE}
combined_total_ethical_rate <- combined |>
  dplyr::select(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`, `Total food and beverage expenditures Year 2`, `Total food and beverage expenditures  Year 1`
  ) |>
  rename(
    `Year 1_E` = `Ethically Produced Rate Year 1`,
    `Year 2_E` = `Ethically Produced Rate Year 2`,
    `Year 1_T` = `Total food and beverage expenditures  Year 1`,
    `Year 2_T` = `Total food and beverage expenditures Year 2`
  ) |>
   mutate(
    `Year 1_T` = as.numeric(`Year 1_T`),
    `Year 2_T` = as.numeric(`Year 2_T`),
    `Year 1_E` = as.numeric(`Year 1_E`),
    `Year 2_E`= as.numeric(`Year 2_E`)
  )

```

```{r, include=FALSE}
library(tidyr)
combined_total_ethical_rate <- combined_total_ethical_rate  |>
  pivot_longer(
    cols = c(`Year 1_E`, `Year 2_E`),
    names_to = "Ethical Year",                        
    values_to = "Ethically Produced Rate"                       
  ) |>
  pivot_longer(
    cols = c(`Year 1_T`, `Year 2_T`),
    names_to = "Budget Year",
    values_to = "Total Budget"
  )
```

```{r, include=FALSE}
total_ethical_cleaned <- combined_total_ethical_rate  |>
  dplyr:: select("Ethically Produced Rate"  , "Total Budget")
  #filter(!is.na("Ethically Produced Rate")) |>
    #filter(!is.na(`Total Budget`)) 
total_ethical_cleaned

```

```{r, include=FALSE}
total_ethical_cleaned  <- total_ethical_cleaned |>
  filter(!is.na(`Ethically Produced Rate`)) |>
    filter(!is.na(`Total Budget`)) 
total_ethical_cleaned

```

```{r fig-regression-ethical, fig.cap="Scatterplot Showing Association Between Budget and Ethical Production Rate", echo=FALSE}
library(ggplot2)
ggplot(
  data = total_ethical_cleaned,
  mapping = aes(
    x = `Total Budget`,
    y = `Ethically Produced Rate`
  )
) + 
  theme_classic()+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color = "red")
```

From @fig-regression-ethical, there is a weak positive association between the two variable.

```{r, include=FALSE}
library(janitor)
cleaned_ethical_rate <- clean_names(total_ethical_cleaned)

glimpse(total_ethical_cleaned)
cleaned_ethical_rate
  
```

The resulting model shows that there is no statistically significant association between the two variable. (t = 1.558 with 98 degrees of freedom, p \>0.05) 

 **Linear equation** :Ethically Produced Rate = 6.825+ 0.0000001672 \* Total Budget

```{r, echo=FALSE}
ethical_budget_model =lm( ethically_produced_rate ~ total_budget, data= cleaned_ethical_rate)
summary(ethical_budget_model)
```

So it can be concluded that there is no statistically significant association between ethically produced rate and total expenditure on food and dining.Any coefficient we get might be artifact due to data limitation.

## Budget Class - Ethically Produced Rate

In this part, we will conduct an ANOVA test to measure any significant differences between the mean ethically produced rate among the three budget classes (@tbl-budgetclass)

- **Null Hypothesis** ($H_0$): All means are equal,
$$\mu_1 = \mu_2 = \mu_3$$
-  **Alternative Hypothesis**($H_1$): At least one budget class mean ethically produced rate is different from others.

```{r, include=FALSE}
class_ethical <- combined |>
  
  dplyr:: select(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`, `Budget Class Year 1`)|>
  mutate("budget_class" = `Budget Class Year 1`)|>
  dplyr::select(budget_class, `Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`) |>
  mutate(`Ethically Produced Rate Year 1`= as.numeric(`Ethically Produced Rate Year 1`))|>
    mutate(`Ethically Produced Rate Year 2`= as.numeric(`Ethically Produced Rate Year 2`))
class_ethical
```

```{r, include=FALSE}
class_ethical <- class_ethical |>
  pivot_longer(
    cols = c(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`),
    names_to = "Year",
    values_to = "value"
  )
  

```

```{r, include=FALSE}
class_ethical_result <- class_ethical |> 
  dplyr:: select(budget_class, value)|>
  filter(!is.na(budget_class))|>
  filter(!is.na(value))|>
  rename(ethically_produced_rate = value)|>
  mutate(budget_class = factor(budget_class))


```

```{r, echo=FALSE}
ethical_summary_1<- class_ethical_result |>
  group_by(budget_class) |>
  summarize(mean_rate = mean(ethically_produced_rate),std.dev_rate = sd(ethically_produced_rate),std.err=sd(ethically_produced_rate)/sqrt(n()))
ethical_summary_1
```

From the summary table, we can see that the p\>0.05 (significance level). So it is not statistically significant.So we fail to reject the null hypothesis stating that all budget class have equal mean ethically produced rate. 

Implying that, the budget classes (lower and higher both) is not significantly associated with ethically produced rate. An institution with lower budget might perform similarly in terms of ethical produce compared to an institution with higher budget. 


```{r, echo= FALSE}
one.way.ethical <- aov(ethically_produced_rate~  budget_class, data = class_ethical_result)
summary(one.way.ethical)
```

```{r, echo=FALSE}
anova(one.way.ethical)
```


## Score ratio over the years

We will analyse how the score ratios have changed between the year 1 and year 2 using a paired t-test method.

First we will check if the data fulfills the conditions for conducting the paired t-test. - The sample size \> 30, so we do not need to check if it follows a normal distribution - The samples are paired since they were collected from measuring twice the score ratio of the same institutions.

- **Null Hypothesis** ($H_0$): $\mu_d = 0$; the mean difference between the paired score ratios is zero.  
- **Alternative Hypothesis** ($H_1$): $\mu_d \neq 0$; the mean difference between the paired score ratios is not zero.

```{r, include=FALSE}
score_years <- combined |>
  dplyr:: select(`Score ratio Year 1`, `Score ratio Year 2`)|>
  filter(!is.na(`Score ratio Year 1`)) |>
  filter(!is.na(`Score ratio Year 2`))
score_years
```

```{r, include=FALSE}
score_year_plot <- score_years |>
  pivot_longer(
    cols = c(`Score ratio Year 1`, `Score ratio Year 2`),
    names_to = "Year",
    values_to = "Score"
  )
score_year_plot
```

Initially, to visualize the distribution of score ratio we use a box plot:

```{r fig-dist, fig.cap = "", echo=FALSE}
ggboxplot(score_year_plot,
          x = "Year", y = "Score", 
          color = "Year",
          palette = c("#00AFBB", "#E7B800"),
          add = "jitter",
          order = c("Score ratio Year 1", "Score ratio Year 2"),
          ylab = "Score", xlab = "Years")+
  theme_classic()+
  coord_flip()
  
```

@fig-dist shows that that the median is almost similar.

```{r, echo=FALSE}
score_year_result <- t.test(  score_years$`Score ratio Year 1`, score_years$`Score ratio Year 2`, paired = TRUE, alternative = "two.sided"
                              )
score_year_result
```
The mean difference (Year 1- Year 2) is 0.0182916; which is not statistically significant. We fail to reject the null hypothesis. Implying that score ratios remained consistent over the years. The institutions had a consistent dining performance over the course of two consecutive reporting years.

## Ethically produced rate over the years

We will analyse how the ethically produced rates have changed between the year 1 and year 2 using a paired t-test method.

First we will check if the data fulfills the conditions for conducting the paired t-test.

- The sample size \< 30, so we do need to check if it follows a normal distribution

```{r, include=FALSE}
ethical_years <- combined |>
  dplyr:: select(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`)|>
  filter(!is.na(`Ethically Produced Rate Year 1`)) |>
  filter(!is.na(`Ethically Produced Rate Year 2`))|>
  mutate(`Ethically Produced Rate Year 1`= as.numeric(`Ethically Produced Rate Year 1`))|>
  mutate(`Ethically Produced Rate Year 2`= as.numeric(`Ethically Produced Rate Year 2`))
ethical_years
```

```{r}
shapiro.test(ethical_years$`Ethically Produced Rate Year 1`)
shapiro.test(ethical_years$`Ethically Produced Rate Year 2`)
```

We can see that they do not follow a normal data distribution. For this, a rank ordered test is better here but it is outside of the study's scope.

## Association between rating status & category of institutions

To compare between the two categorical variable rating status and category, we use chi-squared test of independence.

```{r, include=FALSE}
rate_category<- combined |>
  dplyr:: select(`Rating Status`, Status)|>
  filter(`Rating Status` == c( "Silver"  ,      "Gold", "Bronze", "Platinum"    ))|>
  rename(Category = Status)|>
  group_by(Category, `Rating Status`)|>
  summarize(count = n())



rate_category
```

```{r, include=FALSE}
tab_rate_category<-tibble::tribble(~Category,~Rating,~count,
                           "Flagship", "Gold", 4,
                            "Flagship", "Platinum", 1,
                            "Flagship", "Silver", 3,
                            "Flagship", "Bronze", 0,
                           
                          "Private", "Gold", 4,
                            "Private", "Platinum", 0,
                            "Private", "Silver", 0,
                            "Private", "Bronze", 1,
                           ) |>
  tidyr::pivot_wider(names_from = Rating, values_from=count) |>
  tibble::column_to_rownames(var="Category") |>
  as.matrix() |>
  as.table()
tab_rate_category
```

```{r, echo=FALSE}
tab_rate_category |>
  chisq.test(correct = FALSE)
```

```{r, echo=FALSE}
library(vcd)
assocstats(tab_rate_category)
```

```{r fig-mosaic, fig.cap = "", echo=FALSE}
mosaic(tab_rate_category, gp=shading_Friendly,
labeling=labeling_residuals)
```

Cramer's V is 0.592 which means it has a large effect size. Sample size is small; there is a probable association however given small sample size and too much variability - that causes insignificant p value.

From the @fig-mosaic, we can see for private institutions the most probable rating is Gold and the next likely rating is Bronze. But it is much less likely than Gold. For flagship institutions, the probability of having a gold status is less than expected (if there were no difference between the category of institutions)

## Rating status & Ethically produced rate

In this part, we will conduct an ANOVA test to measure any significant differences between the mean ethically produced rate among the rating statuses (@sec-rating-status)

- **Null Hypothesis** ($H_0$): All rating status pairs have equal mean ethically produced rates:  
  $\mu_{\text{Gold}} = \mu_{\text{Platinum}} = \mu_{\text{Silver}} = \mu_{\text{Not Available}}$
  
- **Alternative Hypothesis** ($H_1$): At least one pair of rating statuses has a different mean ethically produced rate.


The significance level is 0.01 here with a confidence level of 90%

```{r, include=FALSE}
rate_ethical <- combined |>
  dplyr:: select(`Rating Status`, `Ethically Produced Rate Year 1`)|>
rename (rating_status = `Rating Status`,
        ethical_rate = `Ethically Produced Rate Year 1`)|>
  filter(!is.na(ethical_rate))|>
  mutate(rating_status=factor(rating_status))
rate_ethical
```

```{r, echo=FALSE}
one.way.ratethical <- aov(ethical_rate ~ rating_status, data = rate_ethical)
summary(one.way.ratethical)
```

From the results we can see that the p \< 0.01; so it is statistically significant. It can be concluded that at least one of the rating status above has a different mean ethically produced rate.

### Post-Hoc ANOVA Analysis

Now that we know that it is statistically significant, we will conduct Post-Hoc analysis to specify which mean is different. Again, here the significance level is 0.01 with a 90% confidence interval.

```{r, echo=FALSE}
posthoc_analysis <- emmeans(one.way.ratethical, pairwise ~ rating_status,
        adjust = "bonferroni",
        level = 0.90)
summary(posthoc_analysis)


```

When analyzing the p-values, Not Available - Silver has a p value \< 0.1, so at 90% confidence level these two groups have statistically significant different mean ethically produced rate.However it is important to note that at 95% confidence level, the result is not significant.

It means that Not Available-Silver pair has a different statistically significant mean ethically produced rate. Not available groups have a higher mean ethically produced rate than silver status. Noting that, Not available means that the institutions no longer have an active rating(so all of their ratings have expired), there is a possibility of the expired ratings being gold, platinum, silver or bronze.

Including all the expired rating status would help the study more into truly understanding the difference between the gold, platinum, silver or bronze ratings but this is outside the study's scope.

```{r,include=FALSE}
state_avg <- combined |>
  dplyr:: select(`State Name`, `Total food and beverage expenditures  Year 1`) |>
  rename(food_budget = `Total food and beverage expenditures  Year 1`)|>
  rename(state_names = `State Name`)
```

```{r,include=FALSE}
state_avg_grouped <- state_avg |>
group_by(state_names)|>
  summarize(total_food_budget = mean(food_budget, na.arm = TRUE))|>

mutate(total_food_budget = as.numeric(total_food_budget))|>
filter(!is.na(total_food_budget))
state_avg_grouped
```

```{r,include=FALSE}
library(USAboundaries)
states_sf <- us_states()
library(dplyr)
library(sf)
library(USAboundaries)
```

```{r,include=FALSE}
states_sf <- states_sf |>
  mutate(state_abbr = tolower(state_abbr))

state_avg_total <- state_avg_grouped |>
  mutate(state_names = tolower(state_names))
```

```{r,include=FALSE}
merged_state_total <- states_sf |>
  left_join(state_avg_total, by = c("state_abbr" = "state_names"))

```

```{r,include=FALSE}
library(dplyr)
merged_state_total_sf <-merged_state_total |>
  filter(!is.na(total_food_budget)) |>
  
  
mutate(state_name_upper = toupper(state_abbr))|>
  filter(!is.na(state_name_upper))
merged_state_total_sf
```

## Total Food and Beverage Expenditure Across the United States

To visualize which colleges across the states in the US has highest food and beverage expenditure, we will map it by two categories : Flagship, Private.

It is important to note, we are taking their one year expenditure. Since the reporting years are different for each college, we are looking at a range of years. All of the colleges' data hence come from the year 2022-2025.

```{r,include=FALSE}
library(dplyr)
state_avg_faceted <- combined |>
  dplyr:: select(Status,  `Total food and beverage expenditures  Year 1`, `State Name`)|>
  
  rename(food_budget = `Total food and beverage expenditures  Year 1`)|>
  rename(state_names = `State Name`)
state_avg_faceted
```

```{r,include=FALSE}
states_sf <- states_sf |>
  mutate(state_abbr = tolower(state_abbr))

state_avg_total_faceted <- state_avg_faceted |>
  mutate(state_names = tolower(state_names))
```

```{r,include=FALSE}
merged_state_total_faceted <- states_sf |>
  left_join(state_avg_total_faceted, by = c("state_abbr" = "state_names"))

```

```{r, include=FALSE}
merged_state_total_faceted <- merged_state_total_faceted |>
  mutate(state_name_upper = toupper(state_abbr))|>
  filter(!is.na(Status))
merged_state_total_faceted
  
```

```{r,include=FALSE}
map_budget<- tm_shape(merged_state_total_faceted) +
  tm_polygons()
map_budget
```

```{r fig-state-total, fig.cap = "Total Food and Beverage Expenditure Across the United States (2022-2025)", echo=FALSE}

tmap_mode("plot")
map_budget<- tm_shape(merged_state_total_faceted) +
  tm_fill("food_budget",
        style = "quantile",
        palette = "YlGnBu",

        thres.poly = 0)+

  tm_facets(by = "Status", 
            free.coords = TRUE, 
            drop.shapes = TRUE) +
  tm_layout(
    legend.show = TRUE,
            title.position = c("center", "top"), 
            title.size = 2)+
  tm_borders(alpha = 0.5)+
  tm_text(text =  "state_name_upper", size = 0.4,
          remove.overlap = TRUE, 
        fontface = "bold",
        just = "left",
         shadow = TRUE)+
  
tm_style("gray") 
map_budget

```

From the @fig-state-total, the darker region represents higher amount of money spent.

In the flagship institutions:

These are the following states that have higher expenditure: California (CA), Illinois (IL), Indiana (IN), Virginia (VA), Connecticut (CT), Massachusetts (MA), New York (NY)

These are the following states that have lesser expenditure: Montana (MT), Arizona (AZ), Kansas (KS), Iowa (IA), Virginia (VA), New Hampshire (NH)

Florida has the least expenditure.

In the private institutions:

These are the following states that have higher expenditure: Washington (WA), Virginia (VA), Wisconsin (WI), Ohio (OH), Georgia (GA), New Jersey (NJ)

These are the following states that have lesser expenditure: California (CA), Michigan (MI), New York (NY), North Carolina (NC), Maine (ME), New Hampshire (NH)

Now that we know which states have higher expenditure, we see if the same states have similar statistic on ethically produced rate.

## Ethically Produced Rate Comparison Across US States

STARS AASHE has a field for institutions to report their ethically/sustainable produced rate for the institutions {#sec-intro-stars} and {#sec-intro-ceeds}.

### Flagship

The plot in the @fig-ethicalrate-flagship explains the distribution of ethically produced rate across flagship US colleges in different states. The black line represents the average ethically produced rate for all colleges which is around 8% to 9% here. The x axis represents the categorical variable : States and the y axis represents Ethically Produced Rate of foods used in dining. The states below the line produces below average ethical food produce. The states above the line produces more than average ethical food produce.

Predictably so, California, Vermont and Massachusetts have the highest mean rate of ethically produced rate.Followed by Montana, Colorado,North Carolina and Utah. Pennsylvania, Connecticut, Iowa, Texas, Illinois, New York, Louisiana, Maryland, Virginia has lower than average mean rate of ethically produced rate.

```{r, include=FALSE}
library(dplyr)
library(forcats)
```

```{r, include=FALSE}
library(tidyverse)
states_ethical_rate_1 <- combined|>
  dplyr:: select(`State Name`, `Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`, Status)|>

  mutate(`Ethically Produced Rate Year 1` = as.numeric(`Ethically Produced Rate Year 1`))|>
  mutate(`Ethically Produced Rate Year 2` = as.numeric(`Ethically Produced Rate Year 2`))|>
  pivot_longer(cols = c(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`),
               names_to = "Year",
               values_to = "value")|>
  filter(!is.na(value)) |>
  filter(Status == "Flagship")
states_ethical_rate_1
```

```{r, include=FALSE}
states_ethical_rate_2 <- combined|>
  dplyr:: select(`State Name`, `Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`, Status)|>

  mutate(`Ethically Produced Rate Year 1` = as.numeric(`Ethically Produced Rate Year 1`))|>
  mutate(`Ethically Produced Rate Year 2` = as.numeric(`Ethically Produced Rate Year 2`))|>
  pivot_longer(cols = c(`Ethically Produced Rate Year 1`, `Ethically Produced Rate Year 2`),
               names_to = "Year",
               values_to = "value")|>
  filter(!is.na(value)) |>
  filter(Status == "Private")
states_ethical_rate_2
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
library(forcats)
plot_states_ethical_1 <- states_ethical_rate_1 |>
  group_by(`State Name`) |>

  summarize(mean_rate = mean(value, na.rm = TRUE)) |>
  mutate(`State Name` = as.factor(`State Name`)) |>
  mutate(order = fct_reorder(`State Name`, mean_rate))
plot_states_ethical_1
```

```{r, include=FALSE}
plot_states_ethical_2 <- states_ethical_rate_2 |>
  group_by(`State Name`) |>

  summarize(mean_rate = mean(value, na.rm = TRUE)) |>
  mutate(`State Name` = as.factor(`State Name`)) |>
  mutate(order = fct_reorder(`State Name`, mean_rate))
plot_states_ethical_2
```

```{r tbl:private-ethical, include=FALSE}
plot_states_ethical_private<- plot_states_ethical_2|>
  dplyr::select(`State Name`, mean_rate)
plot_states_ethical_private
```

```{r, include=FALSE}
overall_mean_1 <- mean(plot_states_ethical_1$mean_rate, na.rm = TRUE)
overall_mean_2 <-mean(plot_states_ethical_2$mean_rate, na.rm = TRUE)
```

```{r fig-ethicalrate-flagship,fig.cap="", echo=FALSE}

plotted_states_ethical_flagship <- plot_states_ethical_1 |>
  ggplot(aes(x = order, y = mean_rate)) +
  
  geom_hline(yintercept = overall_mean_1,
             color = "black", size = 1) +
  
  geom_segment(aes(x = order,
                   xend = order,
                   y = overall_mean_1,
                   yend = mean_rate),
               color = "black") +
  
  geom_point(aes(color = mean_rate), size = 4) +
  
  scale_color_gradient(low = "yellow", high = "darkblue") +
  
  labs(title = "Ethical Rate Comparison(Flagship)",
       x = "", y = "Rate")+
  
  
 theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 15, face = "bold", color = "black"),
    axis.line = element_line(color = "black"),
plot.caption = element_text(hjust = 0.4),
    axis.ticks = element_line(color = "black"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
  
plotted_states_ethical_flagship
```

### Private

The plot in the @fig-ethicalrate-priv explains the distribution of ethically produced rate across private universities in the US across different states. More colleges fall below average threshold rate of ethically produced rate for all private institutions.Texas,California and Georgia has the highest rate of ethically produced product.

```{r fig-ethicalrate-priv,fig.cap="", echo=FALSE}

plotted_states_ethical_private <- plot_states_ethical_2 |>
  ggplot(aes(x = order, y = mean_rate)) +
  
  geom_hline(yintercept = overall_mean_1,
             color = "black", size = 1) +
  
  geom_segment(aes(x = order,
                   xend = order,
                   y = overall_mean_2,
                   yend = mean_rate),
               color = "black") +
  
  geom_point(aes(color = mean_rate), size = 4) +
  
  scale_color_gradient(low = "yellow", high = "darkblue") +
  
  labs(title = "Ethical Rate Comparison(Private)",
       x = "", y = "Rate")+
  
  
 theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 15, face = "bold", color = "black"),
    axis.line = element_line(color = "black"),
plot.caption = element_text(hjust = 0.4),
    axis.ticks = element_line(color = "black"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
  
plotted_states_ethical_private
```

## Score ratio in Institutes Across the United States

Now, we are going to visualize how the score ratio differs in institutions across states. We calculated the average score ratio over the two available years' for the institutions across states. The grouping variable here is states (which includes both flagship and private institutions). We have data ranging in the year from 2012-2025.

```{r, include=FALSE}
score_ratio_state <- combined |>
 dplyr:: select(`State Name`, `Score ratio Year 1`, `Score ratio Year 2`)|>
  group_by(`State Name`) |>
  


   mutate(across(everything(), ~replace_na(.x, 0))) |>
mutate(avg_ratio = rowMeans(across(c(`Score ratio Year 1`, `Score ratio Year 2`)), na.rm = TRUE))|>
  filter(!is.na(avg_ratio))
  #group_by(`State Name`) |>
  #mutate (total_avg_ratio = mean(avg_ratio))

score_ratio_state
```

```{r, include=FALSE}
total_score_ratio_state <- score_ratio_state |>
  group_by(`State Name`) |>
  mutate(total_avg_ratio = mean(avg_ratio))|>
  distinct(`State Name`, total_avg_ratio, .keep_all = TRUE) |>
  dplyr::select(`State Name`, total_avg_ratio) 



total_score_ratio_state
```

```{r, include=FALSE}
states_sf <- us_states()
```

```{r, include=FALSE}
states_sf <- us_states(resolution = "low")
```

```{r, include=FALSE}
us_states_proj <- st_transform(states_sf, 2163)
```

```{r, include=FALSE}
us_states_proj <- us_states_proj |>
  mutate(state_abbr = tolower(state_abbr))
```

```{r, include=FALSE}
total_score_ratio_state <- total_score_ratio_state |>
  mutate(`State Name` = tolower(`State Name`))|>
rename(state_name = `State Name`)

```

```{r, include=FALSE}
merged_state_score_ratio <- us_states_proj  |>
  left_join(total_score_ratio_state, by = c("state_abbr" = "state_name"))
```

```{r, include=FALSE}
merged_state_score_ratio <- merged_state_score_ratio |>
  filter(state_abbr != "nm")

```

```{r, include=FALSE}
merged_state_score_ratio_ncont <- 
  cartogram_ncont(merged_state_score_ratio,
                  weight = "total_avg_ratio") 
```

```{r, include=FALSE}
label_fun <- function(x) paste0(round(x, 2))

```

```{r, include=FALSE}
state_centroids <- merged_state_score_ratio_ncont|>
  st_centroid() |> 
  mutate(
    lon = st_coordinates(geometry)[,1],
    lat = st_coordinates(geometry)[,2]
  ) |>
  mutate(state_abbr = toupper(state_abbr))
```

```{r fig-ratio-state, fig.cap=" ", echo=FALSE}
m3 <- 
ggplot(merged_state_score_ratio_ncont) +
  geom_sf(aes(fill = total_avg_ratio), color = "white") +
   scale_fill_distiller(
    palette = "YlOrRd",
    direction = 1,   
    labels = label_fun,
    n.breaks = 15,
    guide = guide_colorsteps(
      barwidth = 15,
      barheight = 0.5,
      title = "Average Ratio",
      title.position = "right",
      title.vjust = 0.1)) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, margin = margin(b = 10)),
    legend.position = "top",
    legend.margin = margin(t = 10, b = 2),
    legend.title = element_text(size = 7),
    legend.text = element_text(angle = 45, margin = margin(t = 5))
  ) +
  labs(title = "Sustainably Produced Dining Food Score Ratio in US States(2012-2025)") +
  
geom_text(data = state_centroids, 
          aes(x = lon, y = lat, label = state_abbr)
          , size = 1.6, 
          position = position_jitter(width = 0.1, height = 0.1))

  
m3
```

From the @fig-ratio-state, we can see Georgia, Oklahoma has the highest average score ratio among all of the states.The more the score ratio the darker the region. The lighter (yellow) regions indicate less score ratio and the orange regions indicate medium score ratio compared to the other two parts.

# Analysis

As discussed in @sec-research-smith, some of categories have a higher local sourcing in both years (2017 and 2018) two years. The proportion of a few categories which are local have increased over the course of 2017-2018 compared to the non-local items. It indicates Smith College's effort to promote sustainability in the dining impact area. This is a shift toward ethical and sustainable sourcing practice.We also explored the association between item cost and the likelihood of it being locally sourced using logistic regression. For both the years 2017 and 2018, lower cost were more clustered around an item being non-local. The regression analysis tells us that the model performs well(better in 2018 compared to 2017) predicting an item being local/non-local. In 2023, Smith College's highest spending was allocated to Real food. These spending patterns highlight the college's commitment to support local food sources and sustainable food procurement.

Next, as discussed in @sec-research-consortium, we can see that Hampshire College has the highest sustainable produce percentage and the second highest local produce percentage. According to @hampshire, this college has focused on raising food in a way that is more climate resilient. The Hampshire College Farm is not a commercial firm rather it focuses more on adhering to student experience.This farm produces meat, vegetable and other produces to the college in its 80 acre area.They also partnered with a non profit named Momentum Ag--which solidifies its commitment towards strengthening local producers. Amherst College has the highest locally sourced produce share. According to Amherst @amherst, their Book and Plow Farm produces over 20,000 lbs of fresh produce each year. Their difference in percentage whether it is for local or sustainable may be a result of the availability of college owned farms, larger student body, and proximity to local production places. However, all of the colleges in the consortium show effort in local and sustainable produce

Finally, on the @sec-results, we identified some pattern, trend, similarities and differences between flagship and private institutions.Flagship institutions consistently spent a higher amount of money than the private ones in the same state as seen in @fig-state-total. Given how higher score ratios have a positive correlation with an institutions food and beverage expenditure, higher the budget the higher score ratio. But score ratios stayed consistent for both flagship and private over the two years. Which means dining performance did not significantly differ for both categories in both years. While there were institutions that performed significantly better at times, they were considered outliers for the distribution.However, given their performance in the score ratio has not been statistically different in these two categories, indicating the private institutions are maintaining comparable performance with lesser expenditure. This can be due to the fact that flagship institutions have a larger student population.

After analyzing the total expenditure and ethically produced rate - we found out that they are not associated, yet we recognized a trend at institutions. The average threshold of the mean ethically produced rate for both Flagship (@fig-ethicalrate-flagship) and Private (@fig-ethicalrate-priv) institutions was between 7% and 9%. More colleges fall below the average threshold in private institutions, whereas more colleges fall above the average threshold in flagship institutions. This suggests a minimal difference in ethical sourcing among flagship and private institutions.

When we analyzed the rating statuses across these two categories, most of the private institutions had gold ratings. This is important to note that the ratings are based on overall performance and not just the dining area. If there were truly no difference between the categories, the probability of gold ratings are less than expected for flagship institutions. They had more diverse ratings (given that we took all flagship institutions into account compared to the randomly chosen colleges for each state). But this rating does not have any statistically significant association with ethically produced rate either.

# Limitation

While we were testing out for the changes in variables over the years, not all institutions provided all of the data requested by the STARS framework – that led to inconsistency in tests and might be a reason for statistically insignificant results. Some colleges had expired ratings and no active rating for any year so they were labelled as “not available”. Their lack of active rating might skew the results into making them less significant. We also chose all of the institutions for flagship colleges for each state but only chose one college per state (for private colleges) for the scope of this research project – that leads to data limitation and skewed results in state by state comparison across categories and summary results. It is a self reported tool, so there is a possibility of inconsistent information of data points across institutions. We are doing this comparison over 2 years and the years are not similar across all institutions. For example : some colleges have information about 2022 & 2023, some other colleges have information about 2017 & 2015. The year by year comparison might be biased due to how the socio-economic and geopolitical environment has changed over the years and might not be the same for all.

# Future Scope of Work

This study had a section where we did cross comparison on categories of institutions and years. However, given the analysis was for 2 years hence a longer term analysis (preferably for more than five years) might reveal more significant trends. With the data of all institutions per state, we can have more comprehensive understanding of the averages by each state.
